Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-14 manuscript/manuscript.tex
rewrite of phylobayes section

Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
However, every substitution model has a stationary state, defining the expected sequence composition after a long evolutionary time. 
In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.

@@

We further investigated the effect of ``shifting" preferences by comparing behavior of ExpCM across trees with increasing overall divergence from the focal sequence of the deep mutational scans. 
Based on our results with the large HA trees, we expected the preferences to be more "relevant" on trees which have a low overall sequence divergence from the DMS focal sequence.

We use the value of the ExpCM stringency parameter to measure the ``relevance" of a given DMS preference set and tree. 
A stringency parameter with a value greater than one indicates that selection in nature prefers the same amino acids as selection in lab but with greater strength, or that the preferences are ``relevant" to evolution in nature. 
Conversely, when the stringency parameter is fit to a value close to zero, the preferences are not ``relevant" and de-emphasize the measurements from the lab so that the preferences are flat. 
Using the ExpCM (H1 preferences), we observe an inverse relationship between overall divergence and the stringency. 
As the tree includes more and more sequences which are further from the focal sequence, the relevance of the preferences declines. 
This indicates that the preferences truly are most relevant on trees with sequences closely related to the focal sequence (short branches).

While the preferences are most relevant for the trees with short branches, due to the low divergence from the focal sequence, the effect of site-specific stationary states on branch length estimation is most important on long branches. 
On the low divergence tree, with the most ``relevant" preferences, roughly the same branch length estimated by the GY94+$\Gamma\omega$  and the ExpCM+$\Gamma\omega$. 
In fact, it is not until the high divergence tree that a true effect is seen, in the branches from the focal sequences. 
The effect from site-specific substitution models doesn't really have an effect until the branch lengths are a particular length.

This leads to a regime with competing effects. 
As the branch lengths increase, the ``importance" of having a site-specific stationary state increases. 
At the same time, the ``relevance" of an given site-specific stationary decreases as the divergence increases. 
Current models which have one stationary state across the entire tree, site-specific or not, will not be able to address this tension.

@@

Finally, we compared the branch lengths estimated by ExpCM to branch lengths estimated by another site-specific stationary state model. 
We inferred the branch lengths on the ``high" divergence tree using the mutation-selection model implement in \texttt{phylobayes}. 
The amino-acid composition parameters of the mutation-selection model are estimated in a Bayesian framework in \texttt{phylobayes}, rather than defined \textit{a priori} as in the ExpCM.

Both site-specific stationary state models estimate approximately equivalent branch lengths. 
For a given branch, the two estimates are not exactly the same but each model estimate roughly the same number of branches as longer and to a very similar effect size. 
However, the locality of the ExpCM can still be seen. 
Branch lengths from the focal sequences are enriched in the branches which lengths are estimated to be longer by the ExpCM model. 
That is, the preferences are still more accurate directly around the focal sequence than the amino-acid composition parameters estimated by \texttt{phylobayes} across the entire tree. 
Altogether, this shows that site-specific models, even those which define the stationary state by different strategies, estimate longer branch lengths than uniform stationary state models but that epistasis plays a non-trivial role in how accurate a single stationary state can be.

>>>>

Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
All phylogenetic substitution models \skhcomment{all?} are stochastic matrices and, as such, reach stationary state. 
Stationary state describes the expected sequence composition after a very long evolutionary time and this stationary state sequence composition is independent to time. 
In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.

@@

We looked closer at the effect of shifting preferences on branch length estimation. 
We examined the behavior of the ExpCM on trees with varying levels of overall sequence divergence from the ExpCM focal sequence. 
For each tree, we asked ``does the site-specific stationary state model estimate longer branches than the uniform stationary state model?" and ``is the preference set defining the ExpCM relevant?"

The effect of site-specific stationary state on branch length estimation is seen most strongly on long branches. 
As the full HA tree represents ``high" divergence from the DMS focal sequence, we took subsets the tree to create trees with either ``intermediate" or ``low" divergence from the DMS focal sequence. 
The majority of the branch length estimates are very similar between the GY94+$\Gamma\omega$ and the ExpCM+$\Gamma\omega$ for the ``low" and ``intermediate" alignments. 
Only when the tree has a high overall divergence from the DMS focal sequence, and therefore the longest branches, is there a difference between the two models. 
This result is not unsurprising. 
One of the original motivations for the ``mutation-selection" models was the effect the model would have on long branches specifically.

We wanted to determine the accuracy or relevance of the ExpCM on these subtrees. 
Due to the local branch length extension in \ref{fig:empirical_trees}, we expect that the preferences are most accurate on the low divergence tree. We use the ExpCM stringency parameters as a proxy for accuracy of the preference set. 
The stringency parameter rescales the raw preferences from the deep mutational scan. 
Large stringency parameter values ($>1$) indicate that selection in nature prefers the same amino acids as selection in lab but with a higher stringency.
Or that the preferences measured in lab are relevant for describing the evolution of that protein. 
The stringency parameter for ExpCM(H1 prefs) has the highest value for the low divergence alignment and declines as the overall divergence increases. 
The ExpCM(H3 prefs) show the same pattern. 
The inverse relationship between stringency parameter and divergence supports the hypothesis that epistatic interactions degrade the ExpCM stationary state as defined by DMS preferences. 
\skhcomment{All epistatic? Could there just be a change?}

Current models which have one stationary state across the entire tree, site-specific or not, will not be able to address this tension.

@@

Finally, we compared the branch lengths estimated between two different models with site-specific stationary states. 
We compared the branch lengths estimated by ExpCM+$\Gamma\omega$ (avg) with the branch lengths estimated by the mutation-selection model implemented in \texttt{phylobayes}. 
The \texttt{phylobayes} mutation-selection model estimates the amino-acid preferences which define the stationary state in a Bayesian framework rather than defining them from the DMS \textit{a priori}. 
Branch length estimates are very similar between the ExpCM+$\Gamma\omega$ (avg) and the \texttt{phylobayes} mutation-selection model. 
The equivalence between the two models shows the general importance of site-specific stationary states. 
The local accuracy of the DMS preferences and the ExpCM can still be seen. 
The branch length estimates from either focal sequence (H1 and H3) are longer by the ExpCM than by the \texttt{phylobayes} mutation-selection model. 
This indicates that the DMS preferences are still more accurate for the local branches around the focal sequence than the global preferences estimated by \texttt{phylobayes}.

\skhcomment{What is the average difference in branch length between the two models?}
